\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-05-14 Latex/main.tex
add proposition for transition density of the whole process

%%PACKAGES
\usepackage{authblk}
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{bbold}
\usepackage{float}
\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

@@

@@

The transition density of the velocity Ornstein-Uhlenbeck process is  
\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
	\label{eq: transition RACVM velocity}
\end{equation}

where $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.

%stationary distribution, distribution of the norm, distribution of the phase
This velocity process has stationary distribution $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ \cite{gurarie_correlated_2017}. After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$. In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$ and motivates the interpretation of $\nu$ as the mean velocity norm of the movement.

By integrating the velocity equation over time one gets the following conditional distribution for the location process :
\begin{equation}
	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
	\label{eq: transition RACVM location}
\end{equation}
Here, the exact formula for the covariance matrix $Q(\Delta)$ is the following :

\[
Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
\]
The location process is not markovian, but the whole process $\begin{pmatrix}X(t) &V(t) \end{pmatrix}^\top$ is.
The two components of the location and velocity processes are independent, and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}. Figure~\ref{fig: sample RACVM} shows one sample of a RCVM along with a sample of a constrained version of a RCVM that is defined in section~\ref{section: CRCVM}. Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\

@@

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/illustrative_sample_standard.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard.png}
		\caption{}
	\end{subfigure}
	\caption{(a) Samples of a RACVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $h=5$ min. The red cross indicates the starting point of the process. The samples were obtained from the exact transition densities in equations~\ref{eq: transition RACVM velocity} and~\ref{eq: transition RACVM location}. (b) Samples of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}. Position are sampled from constant time step $h= 1$ min and then downsampled to $5$ min. The samples were obtained from approximate transition densities  with constant covariates on each time step.}
	\label{fig: sample RACVM}
\end{figure}

@@

\subsection{State-space model}
\label{section: state space model}
In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}. This is the approach we use here. In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is 
\[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})= f_{\omega}(\hat{E}^{shore}_{ij},\hat{\Theta}_{ij})\]
 Denote $\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, ,$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
	\frac{1}{\tau_{i}} & -\omega_{ij} \\
	\omega_{ij} & \frac{1}{\tau_{i}}
\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}$.
\begin{comment}
\begin{equation}
	\forall j \in \{1, \cdots, n_{i,pre}\}, \quad 
	\left\{ \begin{array}{l}
		V_{i,j+1}=\exp(-A_{ij}\Delta_{ij})V_{ij}+\zeta_{ij} \\
		X_{i,j+1}=X_{ij}+A_{ij}^{-1}(I_2-\exp(-A_{ij}\Delta_{ij}))V_{ij}+\xi_{ij}
	\end{array}\right. \\
	\label{eq: RACVM autoregressive formula}
\end{equation}

where $\zeta_{ij}$ and $\xi_{ij}$ are centered normal random vectors.
\end{comment}

@@

\begin{equation}
	\forall i \in \{1,\cdots,N\}, \left\{
	\begin{array}{l}
		Y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_i\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}) \\
		\alpha_{i,j+1}=T_{ij} \alpha_{ij} + \eta_{ij}
		\eta_{ij} \sim \mathcal{N}(0,Q_{ij})
	\end{array}
	\right.
	\label{eq: RACVM state space}
\end{equation}
where the link matrices are 
\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_{ij}=\begin{pmatrix} I_2 & A_{ij}^{-1}(I_2-\exp(-A_{ij} \Delta_j)) \\ 0_{2,2} & \exp(-A_{ij} \Delta_j) \end{pmatrix}\]

The vector $\eta_{ij}$ with covariance matrix 
\begin{equation}
	Q_{ij}=\begin{pmatrix}
		M_{ij} & L_{ij} \\
		L_{ij} & N_{ij}
	\end{pmatrix}
	\label{eq: state space covariance}
\end{equation}
where $M_{ij}$, $N_{ij}$ and $L_{ij}$ are block matrices of size $2 \times 2$ such that 
\[M_{ij}=\frac{\sigma_i^2}{C_{ij}}\left( \Delta-2 \frac{\omega_{ij} \sin(\omega_{ij} \Delta_{ij})-\frac{1}{\tau_{i}} \cos(\omega_{ij} \Delta_{ij})}{C_{ij} } \exp\left( -\frac{\Delta_{ij}}{\tau_{ij}} \right) +\frac{\tau_{i}}{2} \left( \frac{\omega_{ij}^2-\frac{3}{\tau_{ij}^2}}{C_{ij}}-\exp\left( -\frac{2\Delta_{ij}}{\tau_i}\right)\right) \right)I_2\]

\[N_{ij}=\frac{\sigma_i^2}{2}\left(1-e^{-\frac{2 \Delta_{ij}}{\tau_{i}}}\right)I_2\]

\[L_{ij}=\begin{pmatrix} a & b \\
	b & a\end{pmatrix}\]
with
\[a=\frac{\sigma_i^2}{2 C_{ij}^2 } \times \left( 1+\exp\left( -\frac{2\Delta_{ij}}{\tau_{i}}\right)-2\exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right)-2\exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right) \cos(\omega_{ij}\Delta_{ij})\right)\]
\[b=\frac{\sigma_{i}^2}{2 C_{ij}^2} \times\left( \exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right) \sin(\omega_{ij} \Delta_{ij})-\frac{\omega_{ij} \tau_{i}}{2} \left(1-\exp\left( -2 \frac{\Delta_{ij}}{\tau_{i}}\right) \right)\right)\]
and $C_{ij}=\frac{1}{\tau_i^2}+\omega_{ij}^2$, $\sigma_{i}=\frac{2\nu_i}{\sqrt{\pi \tau_i}}$.
Again, one can check that for $\omega_{ij}=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.\\

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline
		Coefficient  & True value & Absolute biais  & Relative biais ($\%$) & RMSE\\
		\hline
		$\log(\tau_{pre})$  & 0& 0.08 &  & 0.16\\

		$\log(\nu_{pre})$  & 1.38 & -0.07 & $-5.4 $ & 0.10 \\

		$\omega_0$  &  0.20  & -0.11 & $-58$ & 0.30 \\

		$\omega_1$  & 7.41 & -0.4 & $-5.4 $ & 1.11 \\

		$\omega_2$  & -2.78 & -1.87 &  $67.3$ & 4.30\\

		$\omega_3$  & 4.05 &  0.30 & $7.4$ & 0.96\\

		$\omega_4$  & 14.34 & 0.53 & $3.71$ &  1.62\\

		$\omega_5$  & 4.30 & -4.89 & $-113.7$ &  5.32\\

		$\omega_6$  &  34.38 & -3.07 & $-8.9$ &   3.78\\

		$\omega_7$  & -0.85 & 1.08 & $-126.4 $ & 1.41\\

		$\omega_8$  &  8.81 & -3.91 & $-44.4 $ & 5.88\\

		$\sigma_{\tau,pre}$ & 0.1 & 0.015& $15$ & 0.10\\

		$\sigma_{\nu,pre}$ & 0.3 & 0.007 & $7 $ & 0.04\\
		\hline
	\end{tabular}
	\caption{Measures of estimation performance in the rectangular domain}
	\label{table: estimation performance rect}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline
		Coefficient  & True value  & Absolute biais & Relative biais ($\%$)& RMSE \\
		\hline
		$\log(\tau_{pre})$   & 0     & -0.16 &         & 0.2  \\
		$\log(\nu_{pre}$)     & 1.39  & -0.02 & -1.77     & 0.07 \\
		$\omega_0$   & 0.06  & -0.09 & -148.81   & 0.32 \\
		$\omega_1$ & 11.04 & -1.87 & -16.94    & 2.66 \\
		$\omega_2$ & -3.37 & 1.69  & -50.04    & 2.73 \\
		$\omega_3$ & 7.64  & -2.01 & -26.31    & 2.41 \\
		$\omega_4$ & 23.24 & -2.47 & -10.63    & 4.95 \\
		$\omega_5$ & 7.37  & 1.66  & 22.5      & 5.84 \\
		$\omega_6$ & 41.1  & -4.82 & -11.72    & 5.23 \\
		$\omega_7$ & 2.02  & -0.03 & -1.35     & 1.71 \\
		$\omega_8$ & 10.39 & -1.24 & -11.89    & 2.26 \\
		$\sigma_{\tau,pre}$  & 0.1   & -0.03 & -34.31    & 0.07 \\
		$\sigma_{\nu,pre}$   & 0.1   & -0.06 & -62.63    & 0.07 \\
		\hline
	\end{tabular}
	\caption{Measures of estimation performance in the Scoresby Sound domain}
		\label{table: estimation performance fjords}
\end{table}

@@

\section{Detailed computations of the integrated rotational Ornstein-Uhlenbeck model}
\label{section: detailed RACVM computations}

\subsection{Velocity process}

\subsubsection{Solution of the SDE}

\begin{equation}
	v(t)=\exp(-At) v(0)+ (I_2-\exp(-At))\mu +\frac{2\nu}{\sqrt{\pi \tau}}\int_{0}^{t} \exp(A(s-t)) dW(s)
	\label{eq: RACVM solution}
\end{equation}
In the sequel, we shall use the notation 
\[\zeta(0,t) =\frac{2\nu}{\sqrt{\pi \tau}}\int_{0}^{t} \exp(A(s-t)) dW(s)\]

Replacing $\exp(At)$ by its expression gives more explicitly
\begin{equation}
	v_1(t)=e^{-\frac{t}{\tau}} \left(\cos(\omega t)v_1(0)+\sin(\omega t)v_2(0)\right)+(1-e^{-\frac{t}{\tau}}\cos(\omega t))\mu_1-e^{-\frac{t}{\tau}} \sin(\omega t) \mu_2 +\zeta_1(0,t)
	\label{eq: first random component RACVM}
\end{equation}

\begin{equation}
	v_2(t)=e^{-\frac{t}{\tau}} \left(-\sin(\omega t)v_1(0)+\cos(\omega t)v_2(0)\right)+e^{-\frac{t}{\tau}} \sin(\omega t) \mu_1+(1-e^{-\frac{t}{\tau}}\cos(\omega t))\mu_2 +\zeta_2(0,t)
	\label{eq: second random component RACVM}
\end{equation}
where \begin{align*}
	\zeta_1(0,t)&=\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_1(s) -\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_2(s) \\
	\zeta_2(0,t)&=\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \sin(\omega(s-t)) dW_1(s) +\frac{2\nu}{\sqrt{\pi \tau}} \int_0^t e^{\frac{s-t}{\tau}} \cos(\omega(s-t)) dW_2(s)
\end{align*}
\subsubsection{Autoregressive equation}

For $t \geq 0$ and $\Delta >0$,

\[v(t+\Delta) \vert v(t) \sim \mathcal{N}\left( \mu + \exp(-A\Delta)(v(t)-\mu), \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) \]

Intuitively, the next velocity has a mean that is the sum of the long term mean velocity $\mu$ and the deviation of the previous velocity from this long term mean rotated by an angle $\omega \Delta$.
\subsubsection{Stationary distribution}

>>>>

%%PACKAGES
\usepackage{authblk}
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{bbold}
\usepackage{float}
\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

@@

%% MATHS
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{proposition}{Proposition}[section]

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{remark}
\newtheorem*{example}{Example}

@@

\begin{proposition}
	Let $U(t)=\begin{pmatrix} X(t) & V(t) \end{pmatrix}^\top$ for $t \geq 0$. Then the transition density of the Markov process $U$ is given by 
	\begin{equation}
		U(t+\Delta) \vert U(t)=u \sim \mathcal{N}\left( T(\Delta) u +B(\Delta)\mu, Q(\Delta)\right) \mbox{ for all } \Delta>0 \mbox{ and } u \in \R^4
		\label{eq: X V distribution}
	\end{equation}
where \begin{equation}
	T(\Delta)=\begin{pmatrix} I_2 & A^{-1}(I_2-\exp(-A\Delta)) \\
	0_2 & \exp(-A\Delta) \end{pmatrix} \mbox{ and } B(\Delta)=\begin{pmatrix}
	\Delta I_2 -A^{-1}(I_2-\exp(-A\Delta))\\
	I_2-\exp(-A\Delta)\end{pmatrix}
	\label{eq: link matrices}
	\end{equation}
and the covariance block matrix is given by
\begin{equation}
	Q(\Delta)=\begin{pmatrix}
		q_1(\Delta)I_2 & \Gamma(\Delta) \\
		\Gamma(\Delta)^\top & q_2(\Delta)I_2
	\end{pmatrix}
	\label{eq: covariance matrix}
\end{equation}
with 
\begin{equation}q_1(\Delta)=\frac{\sigma^2}{C}\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{C_{ij} } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{C}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
\end{equation}

\begin{equation}
	q_2(\Delta)=\frac{2\nu^2}{\pi}\left(1-e^{-\frac{2 \Delta}{\tau}}\right)
\end{equation}
\begin{equation}\Gamma(\Delta)=\begin{pmatrix} \gamma_1 & \gamma_2 \\
	-\gamma_2 & \gamma_1\end{pmatrix}
\end{equation}
where
\begin{equation}\gamma_1=\frac{\sigma^2}{2 C } \times \left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega\Delta)\right)
\end{equation}
\begin{equation}\gamma_2=\frac{\sigma^2}{C} \times\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right)
\end{equation}
and we denoted $C=\frac{1}{\tau^2}+\omega^2$ and $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$.
\label{prop: transition density}
\end{proposition}

The formulas derived in \cite{johnson_continuoustime_2008} are a corollary of ~\ref{prop: transition density}, obtained with $\omega=0$. The proof of this proposition is detailed in Appendix.

Equation~\ref{eq: covariance matrix} shows that the two components of the location and velocity processes are independent.
Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
Intuitively, ~\ref{eq: X V distribution} means that the next velocity $V(t+\Delta)$ is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $V(t)$ rotated by an angle $-\omega \Delta$.
 Figure~\ref{fig: sample RACVM} shows one sample of a RCVM along with a sample of a constrained version of a RCVM that is defined in section~\ref{section: CRCVM}.\\

@@

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/illustrative_sample_standard.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard.png}
		\caption{}
	\end{subfigure}
	\caption{(a) Samples of a RCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min. The red cross indicates the starting point of the process. The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}. (b) Samples of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}. Position are sampled from constant time step $\Delta= 1$ min and then downsampled to $5$ min. The samples were obtained from approximate transition densities  with constant covariates on each time step.}
	\label{fig: sample RACVM}
\end{figure}

@@

\subsection{State-space model}
\label{section: state space model}
In most applications, the error between the observation $Y$ and the true position $X$ is not negligible, so that it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}. This is the approach we use here. In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is 
\[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})= f_{\omega}(\hat{E}^{shore}_{ij},\hat{\Theta}_{ij})\]
 Denote $\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, ,$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
	\frac{1}{\tau_{i}} & -\omega_{ij} \\
	\omega_{ij} & \frac{1}{\tau_{i}}
\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}$.

@@

\begin{equation}
	\forall i \in \{1,\cdots,N\}, \left\{
	\begin{array}{l}
		Y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_i\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}) \\
		\alpha_{i,j+1}=T_{ij} \alpha_{ij} + \eta_{ij} \\
		\eta_{ij} \sim \mathcal{N}(0,Q_{ij})
	\end{array}
	\right.
	\label{eq: RACVM state space}
\end{equation}
where $Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix}$, the link matrices $T_{ij}$ and $B_{ij}$ are computed according to~\ref{eq: link matrices} and the covariance matrix is computed according to~\ref{eq: covariance matrix}.
\\

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline
		Coefficient  & True value  & Biais & Relative biais ($\%$) & RMSE \\
		\hline
		 $\log(\tau_0)$   & 0     & 0.03  &         & 0.11 \\
		$\log(\nu_0)$     & 1.39  & -0.02 & -1.55     & 0.06 \\
		$\omega_0$   & 0     & -0.02 &    & 0.22 \\
		$\omega_1$ & 8.03  & -1.29 & -16.05    & 1.64 \\
		$\omega_2$ & -3.14 & 1.55  & -49.39    & 4.13 \\
		$\omega_3$ & 5.08  & 0.21  & 4.05      & 1.23 \\
		$\omega_4$ & 16.8  & -1.86 & -11.08    & 2.32 \\
		$\omega_5$ & 5.03  & -6.03 & -119.78   & 6.39 \\
		$\omega_6$ & 22.52 & -2.83 & -12.56    & 3.27 \\
		$\omega_7$ & 2.2   & -0.14 & -6.21     & 0.93 \\
		$\omega_8$ & 6.17  & -4.57 & -74.09    & 6.14 \\
		$\sigma_{\tau,pre}$  & 0.1   & 0.02  & 21.91     & 0.11 \\
		$\sigma_{\nu,pre}$    & 0.1   & 0.02  & 18.53     & 0.05 \\
		\hline
	\end{tabular}
	\caption{Measures of estimation performance in the fjords for the first $12$h-long trajectories.}
\label{table: estimation performance fjords 12h}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline
		Coefficient   & True value  & Biais & Relative biais ($\%$) & RMSE \\
		\hline
		$\log(\tau_0)$   & 0     & 0.03  &         & 0.09 \\
		$\log(\nu_0)$  & 1.39  & -0.02 & -1.69     & 0.05 \\
		$\omega_0$     & 0     & -0.02 &   & 0.13 \\
		$\omega_1$ & 8.03  & -0.97 & -12.1     & 1.25 \\
		$\omega_2$ & -3.14 & -0.71 & 22.71     & 2.64 \\
		$\omega_3$ & 5.08  & 0.35  & 6.84      & 1.01 \\
		$\omega_4$ & 16.8  & -1.16 & -6.93     & 1.66 \\
		$\omega_5$ & 5.03  & -6.22 & -123.63   & 6.41 \\
		$\omega_6$ & 22.52 & -2.04 & -9.06     & 2.35 \\
		$\omega_7$ & 2.2   & 0.01  & 0.65      & 0.64 \\
		$\omega_8$ & 6.17  & -2.61 & -42.31    & 3.54 \\
		$\sigma_{\tau,pre}$   & 0.1   & 0.01  & 13.05     & 0.08 \\
		$\sigma_{\nu,pre}$    & 0.1   & 0.03  & 29.72     & 0.04 \\
		\hline
	\end{tabular}
	\caption{Measures of estimation performance in the fjords for the $24$h-long trajectories.}
\label{table: estimation performance fjords 24h}
\end{table}

@@

\section{Proof of proposition~\ref{prop: transition density}}
\label{section: transition density proof}

\begin{proof}
The velocity process is an Ornstein-Uhlenbeck process. For $t \geq 0$ and $\Delta >0$,
\begin{equation}
	V(t+\Delta)=\exp(-A\Delta) V(t)+ (I_2-\exp(-A\Delta))\mu +\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) dW(s)
	\label{eq: RACVM solution}
\end{equation}

It has gaussian transition density with mean 
\begin{equation}
	\E(V(t+\Delta \vert V(t)))=\exp(-A\Delta) V(t)+ (I_2-\exp(-A\Delta))\mu 
\end{equation}
and covariance matrix 
\begin{align*}Var(V(t+\Delta) \vert V(t))&=\frac{4\nu^2}{\pi \tau} \int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \exp(A(s-(t+\Delta)))^{\top}\\
	&=\frac{4\nu^2}{\pi \tau} \int_{0}^{\Delta} \exp(-Au) \exp(-Au)^{\top} du 
\end{align*}
Since $\exp(-Au)=\exp(-\frac{u}{\tau}) R_{-\omega u}$, the matrix product  $\exp(-Au) \exp(-Au)^{\top}$ is simply $\exp(-\frac{2u}{\tau}) I_2$. We deduce that the two components of the velocity are independent and have the same variance, written $q_2(\Delta)$ to follow the notations in~\ref{prop: transition density}. The variance is 
\begin{equation}
	q_2(\Delta)=\frac{2\nu^2}{\pi}\left(1-\exp\left(-\frac{2\Delta}{\tau}\right)\right)
\end{equation}
These results are found in \cite{gurarie_correlated_2017}.
In the sequel, we shall use the notation 
\[\zeta(t,s) =\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{s} \exp(A(u-s)) dW(u)\]

\begin{align*}
	X(t+\Delta)&=X(t)+\int_t^{t+\Delta} V(s) ds \mbox{ where } V(s)=\mu+\exp(-A(s-t))(V(t)-\mu)+\zeta(t,s) \\
	&= X(t)+\mu \Delta+\int_t^{t+\Delta} \exp(-A(s-t))(X(t)-\mu) ds +\int_t^{t+\Delta} \zeta(t,s) ds \\
	&= X(t)+\mu \Delta +\int_0^{\Delta} \exp(-As)(X(t)-\mu) ds +\int_t^{t+\Delta} \zeta(t,s) ds \\
	&=X(t)+\mu \Delta + (A^{-1}(V(t)-\mu)-A^{-1}\exp(-A\Delta)(X(t)-\mu))+\int_t^{t+\Delta} \zeta(t,s) ds \\
\end{align*}
Thus, 
\begin{equation}
	X(t+\Delta)=X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu) +\xi(t,t+\Delta) 
\end{equation}
where $ \xi(t,t+\Delta)=\int_t^{t+\Delta} \zeta(t,s)ds$.
The location process is also gaussian with mean 
\begin{equation}
	\E(X(t+\Delta) \vert V(t),X(t)) =X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu)
\end{equation}
To get an expression of the covariance matrix, first rewrite

\begin{align*}
	\xi(t,t+\Delta)&=\int_t^{t+÷\Delta} \frac{2 \nu}{\sqrt{\pi \tau}} \left( \int_t^s \exp(-A((u-s))dW(u) \right) ds \\
	&=\frac{2\nu}{\sqrt{\pi\tau}} \int_t^{t+\Delta} \left( \int_u^{t+\Delta} \exp(A(u-s)) ds \right) dW(u)  \\
	&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} (A^{-1}-A^{-1}\exp(A(u-t-\Delta))) dW(u) \\
	&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) dW(u)
\end{align*}

Then use Ito's isometry
\begin{align*}
	Var(X(t+\Delta)\vert X(t),V(t))&=\frac{4 \nu^2}{\pi \tau} \int_{t}^{t+\Delta} A^{-1} (I_2-\exp(A(u-t-\Delta)))(I_2-\exp(A(u-t-\Delta)))^\top A^{-\top} du \\
	&=\frac{4 \nu^2}{\pi \tau} \int_{0}^{\Delta} A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top} dr
\end{align*}
Now this integral can be computed explicitly since 
\[A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top}=\frac{1}{C} \begin{pmatrix}  f(r) & 0 \\ 0 & f(r)\end{pmatrix}  \]
where $f(r)=1-2\exp\left( -\frac{r}{\tau}\right)\cos(\omega r)+\exp\left(\frac{-2r}{\tau}\right)$ and $C=\frac{1}{\tau^2}+\omega^2$.
We obtain that $X_1(t+\Delta)$ and $X_2(t+\Delta)$ are independent and have the same variance, denoted $q_1(\Delta)$.
Writing $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$, the variance is
\begin{align*}q_1(\Delta)=\frac{\sigma^2}{C} \left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right)
\end{align*}
